---
title: "plant to music script"
subtitle: last updated SZhang 8/20/2022
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    number_sections: yes
---

# Introduction

## pseudocode

1. READ fileS into dataset

   read the C2-C6 Hertz range list as a dataset called NOTEKEY
   
   read your data as a dataset called yourDATA
   
2. RENAME your data column of interest in yourDATA so you don't have to go type it in every where after

3. check yourDATA to see if there's missing values, it's suitable, or if it needs to be transformed using IF statements, e.g. :

IF

    min value of yourDATA < hertz for C2
    
    PRINT "your smallest data value may be too small"
    
ELSE

    PRINT "your smallest value should be fine"
    
etc etc. In yourDATA need to check for:

  A. min & max of y value
  
  B. range of y value
  
  C. stdev of y value [<<optional]
  
4. Turn y value of yourDATA to notes

INITIALIZE i for the NOTEKEY counting, set i = 1

INITIALIZE j for yourDATA counting, set j = 1

INITIALIZE count for the number of y values in yourDATA

CREATE new column 'newnote' for converted notes then

start loop #1

  WHILE j <= count
  
start loop #2 inside loop #1

    WHILE hertz at row i < convert y-value of yourDATA at row j
    
      ADD 1 to i
      
    OTHERWISE
    
      IF converted y-value of yourDATA at j == 0
      
        SET newnote in row j of yourDATA = NA
        
        RESET i to 1
        
        ADD 1 to j
        
        EXIT if-else
        
      ELSE 
      
        SET value of newnote in row j of yourDATA 
        
        = musical note value of row i-1 in NOTEKEY
        
        RESET i to 1
        
        ADD 1 to j
        
        EXIT if-else
        
    EXIT loop #2
    
  OTHERWISE EXIT loop #1
  
  
5. PRINT newnote column to see if it all looks okay

6. EXPORT as new .csv file for use in music apps

## data requirements

Your data should be in column format (like the usual case in R). It should be in .csv format and should contain at least 1 column of numerical data with a header. The header is the name of that column. E.g.

|rootLength|
|----------|
|10.2      |
|12.5      |
|9.7       |
|11.1      |

## additional notes

Your data file and the notekey.csv file should be saved in the same folder that you save this .Rmd file in.
    
# loading data

## setting NOTEKEY
```{r}
#DO NO CHANGE ANYTHING IN THIS CHUNK

NOTEKEY <- read.csv('NOTEKEY.csv')
NOTEKEY$octave <- as.character(NOTEKEY$octave)

#summary(NOTEKEY) #used to check loading, ignore this line
# str(NOTEKEY$note_octave) #used to check loading, ignore this line

#get the hertz value for middle c to use to check again your data range later
note_split<-split(NOTEKEY, NOTEKEY$octave)
middleC <- note_split$'4'

```

## load TESTDATA

### this is start of the part that you can edit

```{r}
#CHANGE STUFF HERE

#instead of exampleData.csv put name of your .csv
#make sure your .csv is located in the same folder where you save this .Rmd!
#blank cells WILL give you errors
yourDATA <- read.csv('kernel.csv') 

#instead of 'perimeter' put the name of your column
colnames(yourDATA)[which(names(yourDATA) == 'height')] <- 'colOfInterest'

#check to see if it all read correctly
summary(yourDATA)
```
# start data processing

## data check

### set values
```{r}
#DO NOT CHANGE ANYTHING IN THIS CHUNK

#hertz range we have to work with
notekey_min = min(NOTEKEY$frequency)
notekey_max = max(NOTEKEY$frequency)
note_stdev = sd(middleC$frequency)

#the number of datapoints in yourDATA
data_count = length(yourDATA$colOfInterest)

```


### check to see if you've missing data
```{r}
#DO NOT CHANGE ANYTHING IN THIS CHUNK

#counter
k = 1

#check to see if you've an NA
while(k <= data_count){
  
 if (is.na(yourDATA$colOfInterest[k]) == FALSE){
   k = k +1
 } else {
   print("You have a missing value in your dependent variable column, consider deleting that row of data.")
   k = data_count + 1
 }
  
}

```
If you received a missing value error message here, go back and delete the row with the missing value, save your .csv file and then go re-read in your data again. No message means you're fine and can move on to the next chunk.

### check to see if your data needs to be transformed
```{r}
#check to see if your data needs to be transformed
#DO NOT CHANGE STUFF IN THIS CHUNK

#your initial data range
your_min = min(yourDATA$colOfInterest) 
your_max = max(yourDATA$colOfInterest) 
your_stdev = sd(yourDATA$colOfInterest)

#this bit is for if your data includes 0 as part of the value
#counter 
l = 1


#sort your data column from smallest to largest by
# 1) initialize new column "colSorted"
yourDATA$colSorted = yourDATA$colOfInterest
# 2) sort data from smallest to largest
yourDATA$colSorted = sort(yourDATA$colSorted)

while(l <= data_count){
  
 if (yourDATA$colSorted[l] == 0){
   l = l +1
 } else {
   your_min = yourDATA$colSorted[l]
   #print(your_min) #this part is for troubleshooting the loop; ignore
   l = data_count + 1
 }
  
}

print('The acceptable numerical range for conversion music is:')
print(summary(NOTEKEY$frequency))

print('Your current data summary is:')
print(summary(yourDATA$colOfInterest))

if (your_min < notekey_min) {
  print ("Your lowest value is too small, its converted frequency will be too low.")
} else if (your_max > notekey_max){
  print ("Your highest value is too big, its converted frequency will be too high.")
}  else {
    print ("The min and max value of your data should be fine.")
  }
    
note_range = notekey_max - notekey_min
your_range = your_max - your_min

if (your_range > note_range){
  print("The range of your data is too wide for this program.")
} else {
  print("The range of your data should be fine.")
}
 
if (your_stdev < note_stdev){
  print("OPTIONAL: the variation in your data is a little on the small side.")
} else {
  print ("The variation in your data should be fine.")
}

```


**If your minimum value is too small**
Consider multiplying a constant value to all data in your column. You want the minimum to be over 66 Hertz. In R you can just type '+ 5'to add 5 to a number. 

**If your maximum value is too big**
Consider dividing a constant value from all data in your column. You want the maximum value to be under 1975 Hertz. In R you can just type '-5' to minus 5 from a number. 

**why multiply and divide**
Multiplication and division is recommended over adding and subtraction in case the data set contains 0 as part of the true values.

**If the range of your data is too wide**
You can either take the square root of all data in your column or divide all data by a constant value, or taking their log. In R you can do square root using 'sqrt()' (you use these functions with the parentheses the same way you would use the average() function in Excel). Divide is just '/', log is 'log()'. (You can also do 'log(x, base = y)', where x is what you want to take log of and y is your base number. In otherwords, if you don't give the base and just type 'log()' the default base is 10.)

**OPTIONAL: If the variation in your data is too small**
There are many ways you can deal with this. You can try subtracting your minimum value from all data, then adding in 0.01*minimum value (to ensure you don't get zeros), then multiply everything by a constant value (to make everything in range -- your are effectively magnifying the variation in your data this way). This part is up to personal taste to be honest.

## how to transform your data (if you need to)
```{r}
#CHANGE THINGS HERE IF YOUR DATA NEEDS TO BE TRANSFORMED

#DIRECTIONS

#1. remove '#' from the front of the transformation you want (line 140-144)
#2. change the value to be added, subtracted, etc to whatever you want
#note adding 


#transformed_add = yourDATA$colOfInterest + 20
transformed_multiply = yourDATA$colOfInterest * 100
#transformed_sqrt = sqrt(yourDATA$colOfInterest)
#transformed_div = yourDATA$colOfInterest / 4
#transformed_log = log(yourDATA$colOfInterest, base = 2)
#biggerVariance = (yourDATA$colOfInterest-your_min+0.01*your_min)*12

#3. remove '#' from the line of code below (line 286) and change the right side of <- to 'transformed_add'or transformed_div' or whatever you chose from above

yourDATA$colOfInterest <- transformed_multiply


#4. re-run the code for the "does your data need to be transformed" section in line 155
#5. repeat steps 1-4 above until you're told your data's all ok. 

#Each time you repeat steps 1-4 remember to:
#A. put '#' back in front of the line of code you used the last time  
#B. delete the '#' out of the line of code you're using this time
#C. change the right side of the <- on line 291 to whichever transformation you're using this time!

#NOTE: you can always reset all transformation by re-reading in your data from your .csv file

```
Do not proceed until the next part until you get the confirmation that your min and max values are okay.

Remember that this step above is iterative! Meaning if the first time you run it you did multiply by 10, then you run it for a second time multiplying by 10, the value you're working with is the original value multiplied by 100!

## turning your numbers to notes
```{r}
#DO NOT CHANGE ANYTHING IN THIS CHUNK

#initializing new column for new notes
newnote = yourDATA$colOfInterest


#initializing, cont: add new column to data set
yourDATA$newnote <- newnote

#integer for NOTEKEY
i = 1
# integer for yourDATA
j = 1

#loop through all the numbers in yourDATA
while(j <= data_count){
  
   #loop through all frequencies in NOTEKEY
 while (yourDATA$colOfInterest[j] > NOTEKEY$frequency[i]){
   i = i +1
 } 
  
  if (yourDATA$newnote[j] == 0){
    yourDATA$newnote[j] = NA
    j = j + 1
    i = 1
  } else{
  
  yourDATA$newnote[j] = NOTEKEY$note_octave[i-1]
  j= j + 1
  i = 1
 }
    
}

```

## output to .csv
```{r}
#DO NOT CHANGE ANYTHING IN THIS CHUNK

#print your new notes to see if it all looks ok
print(yourDATA$newnote)

#export to csv
# a new file will show up in this folder called 'yournotes.csv' 
write.csv(yourDATA, "yournotes.csv", row.names = FALSE)
```

## final notes [pun by accident]
 
You can copy your new column of music notes into many of the free apps available either for your smartphone, computer, or through a browser. Or, if you're inclined, try to play it on your musical instrument of choice. 

For smartphone Bandlab is available for both android and iOS smartphones, 

Typatone is through the web but also available as phone app. Note: it reads the letter, the symbol (#) and the number separately (and can sound very cute, but also limits its range).

NOTE: In the yournotes.csv file, getting NA in the newnote column means that that beat should be silent. (This corresponds to a value of 0 in your dependent variable.)
